---
title: "Diabetes in Pima Indians"
author: "Katherine Rodriguez"
date: "2026-07-10"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\newpage

# Summary
This script uses the Pima Indians diabetes dataset to build predictive models 
for diabetes status. After cleaning impossible values 
(e.g., 0 for glucose, blood pressure, BMI) and recoding the outcome,
we fit a logistic regression and a random forest classifier using tidymodels. 
We compare performance on a train/test split using accuracy and ROC AUC, 
visualize the ROC curve, and examine variable importance to understand which
health metrics are most strongly associated with diabetes.

# Load and prepare data

## Load libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
```

## Load the data
```{r}
diabetes <- read_csv("diabetes.csv")
```

## Lets take a look at the data
```{r}
head(diabetes)
```

# Tidy the data

## Replace values of zeros with NAs for BMI, glucose, and blood pressure recode outcome
```{r}
diabetes <- diabetes |>
  mutate_at(c('bmi', 'glucose', 'blood_pressure'), ~na_if(., 0))|>
  mutate_if(is.character, as.factor) |>
  mutate(
    outcome = recode_factor(outcome, "0" = "No diabetes", "1" = "Diabetes")
  )
```

# Train and test split
```{r}
set.seed(630)
split_diab <- initial_split(diabetes, prop = 0.75, strata = outcome)
train_diab <- training(split_diab)
test_diab <- testing(split_diab)
```

# Recipe: impute, dummy, and normalize
```{r}
rec_diab <- recipe(outcome ~ pregnancies + glucose + blood_pressure + bmi + age + insulin,
                   data = train_diab) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
```

# Model Building

## Logistic Regression Model
```{r}
logit_mod <- logistic_reg(mode = "classification") |>
  set_engine("glm")

logit_wf <- workflow() |>
  add_recipe(rec_diab) |>
  add_model(logit_mod)

logit_fit <- logit_wf |>
  fit(data = train_diab)
```

## Random Forest Model
```{r}
rf_mod <- rand_forest(mode = "classification", trees = 500) |>
  set_engine("ranger", importance = "permutation")

rf_wf <- workflow() |>
  add_recipe(rec_diab) |>
  add_model(rf_mod)

rf_fit <- rf_wf |>
  fit(data = train_diab)
```

# Predictions on test set
```{r}
logit_pred <- predict(logit_fit, new_data = test_diab, type = "prob")|>
  bind_cols(predict(logit_fit, new_data = test_diab, type = "class")) |>
  bind_cols(test_diab |> select(outcome)) |>
  rename(.pred_class = .pred_class)

rf_pred <- predict(rf_fit, new_data = test_diab, type = "prob") |>
  bind_cols(predict(rf_fit, new_data = test_diab, type = "class")) |>
  bind_cols(test_diab |> select(outcome)) |>
  rename(.pred_class = .pred_class)
```

# Metrics: Accuracy and ROC AUC
```{r}
metrics_logit <- tibble(
  model = "Logistic regression",
  accuracy = accuracy(logit_pred, truth = outcome, estimate = .pred_class)$.estimate,
  roc_auc = roc_auc(logit_pred, truth = outcome, .pred_Diabetes, event_level = "second")$.estimate
  
)

metrics_rf <- tibble(
  model = "Random forest",
  accuracy = accuracy(rf_pred, truth = outcome, estimate = .pred_class)$.estimate,
  roc_auc = roc_auc(rf_pred, truth = outcome, .pred_Diabetes, event_level = "second")$.estimate
  
)

bind_rows(metrics_logit, metrics_rf)  

```

**Discussion**

Both models perform well, with exact accuracy of 0.832.
ROC AUC values are slightly different with logistic regression at 0.883,
and random forest at 0.874

# ROC Curve and Variable Importance for Random Forest
```{r}
roc_plot <- rf_pred |>
roc_curve(truth = outcome, .pred_Diabetes, event_level = "second") |>
  autoplot() +
  labs(
    title = "Random Forest ROC Curve (Diabetes)",
    x = "1-Specificity",
    y = "Sensitivity"
  ) 

#Save ROC curve
ggsave("roc_curve.png", roc_plot, width = 6, height = 4)

#Variable importance (random forest)
rf_fitb <- extract_fit_parsnip(rf_fit)$fit
imp <- tibble(
  variable = names(rf_fitb$variable.importance),
  importance = as.numeric(rf_fitb$variable.importance)
) |>
  arrange(desc(importance))
```

# Visualization
```{r}
varimp_plot <- ggplot(imp, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Random Forest Variable Importance",
    x = NULL,
    y = "Importance"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5))

#Save variable importance chart
ggsave("variable_importance.png", varimp_plot, width = 6, height = 4)
```  
  
**Discussion**

As the metrics revealed, both the logistic regression and random forest
models performed well with accuracy both 0.832, while logistic regression
had a slightly higher ROC AUC value. The variable importance chart reveals 
glucose is the strongest predictor of diabetes, consistent with clinical 
explanations of diabetes and what is likely to cause a higher diabetes risk.